---
title: "Tzu-Chun's R coding Exercise"
output: 
  html_document:
    toc: FALSE
---

Hello! For this week's exercise, I will explore the *gapminder* dataset that contains health and income outcomes from 184 countries between 1960 and 2016. This exercise will further evaluate any association between life expectancy and infant mortality or population size. 

First, let's load all R packages that will be used for this exercise
```{r, warning = F, message = F}
## load R packages
library(tidyverse) # tidyverse offers great tools for data wrangling and visualization
library(scales) # to access break formatting functions
library(dslabs) # to use the gapminder data from this package
```

I then loaded the *gapminder* data and checked the data strucutre, summary and class 
```{r}
#Use help to access to the documentation page for gapminder data
help(gapminder)
#get an overview of data structure
str(gapminder)
```

```{r}
#Here is the data summary
summary(gapminder)
```

```{r}
#determine the type of object gapminder is
class(gapminder)
```

## Processing data
Before working on some data visualization and fitting model. I ran a couple lines of code to create new objects and check their data structure and summary. First of all, I use *subset* function to select only African countries by specifying where the continent variable is "Africa", and named the object *africadata*. Next, I ran *str()* and *summary()* to see the structure and summary of both objects. 
```{r}
#assign African countries to a new object
africadata <- subset(gapminder, continent == "Africa")

#create two new objects
imr.le <- africadata[,c("infant_mortality","life_expectancy")]#first one contains "infant_mortality" and "life_expectancy"
pop.le <- africadata[,c("population","life_expectancy")]#first one contains "population " and "life_expectancy"
```

```{r}
#get an overview of the data
str(africadata)
```


```{r}
#check the summary of the data
summary(africadata)
```


```{r}
#take a look at data structure and summary for the new object imr.le
str(imr.le)
summary(imr.le)
```



```{r}
#take a look at data structure and summary for the new object pop.le
str(pop.le)
summary(pop.le)
```

## Plotting
I used *ggplot()* to create two separate scatter plots from the objects I just created, The first one plotted life expectancy against infant mortality, while the second one plotted life expectancy against population size. 
```{r}
#scatter plot of infant mortality and life expectancy from 1960 to 2016
imr.le %>% 
  ggplot(aes(x=infant_mortality, y=life_expectancy)) +
  geom_point() +
  #add plot title
  ggtitle("Relationship between infant mortality and life expectancy for African countries\n from 1960 to 2016") + 
  #change x-axis label
  xlab("Infant mortality (death per 1000)") + 
  #change y-axis label
  ylab("Life expectancy (years)") 
```

The plot shows that there was a negative correlation between infant mortality and life expectancy. 

```{r}
#scatterplot of population and life expectancy from 1960 to 2016
pop.le %>% 
  ggplot(aes(x=population, y=life_expectancy)) +
  geom_point() +
  #transform and reformat x-axis to log scale
  scale_x_continuous(trans = 'log10',
                     labels = trans_format("log10", math_format(10^.x))) + 
  #add plot title
  ggtitle("Relationship between population size and life expectancy for African countries\n from 1960 to 2016") + 
  #change x-axis label
  xlab("Population (log scale)") + 
  #change y-axis label
  ylab("Life expectancy (years)") 
  
```

We can see that there were some streaks following a positive and linear trend between population and life expectancy. This could result from the source that mixed data from different countries across multiple years, so my next step is to try to avoid this noise and only look into the data from a particular year.

## More data processing and plotting
I grouped the data by year and summarized how many missing data in infant mortality across each year, then I filtered out only those years with missing data.
```{r}
#write code that figures out which years have missing data for infant mortality
africadata %>% 
      group_by(year) %>% 
      summarise(infant_mortality_na = sum(is.na(infant_mortality))) %>% 
      filter(infant_mortality_na > 0) #missing up to 1981 and 2016
```
There were some missing data in infant mortality from 1960 to 1981 and 2016 (n=226). Hence, I will only use the data from 2000 to create the same two plots again. 

```{r}
#create new object to extract data for 2000
africadata2000 <- africadata[africadata$year == 2000,]

#take a look at data structure and summary
str(africadata2000)
summary(africadata2000)
```

```{r}
#scatterplot of infant mortality and life expectancy only for year of 2000
africadata2000 %>% 
  ggplot(aes(x=infant_mortality, y=life_expectancy)) +
  geom_point() +
  ggtitle("Relationship between infant mortality and life expectancy for African countries\n in 2000") + #add plot title
  xlab("Infant mortality (death per 1000)") + #change x-axis label
  ylab("Life expectancy (years)") #change y-axis label
```

We still observed a negative correlation between infant mortality and life expectancy from the data in 2000.

```{r}
#scatterplot of population and life expectancy only for year of 2000
africadata2000 %>% 
  ggplot(aes(x=population, y=life_expectancy)) +
  geom_point() +
  scale_x_continuous(trans = 'log10', breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x))) + #transform and reformat x-axis to log scale
  ggtitle("Relationship between population size and life expectancy for African countries\n in 2000") + #add plot title
  xlab("Population (log scale)") + #change x-axis label
  ylab("Life expectancy (years)") #change y-axis label
```

However, we no longer saw any apparent correlation between population size and life expectancy when we looked into only the data in 2000.


## Model fit
The final step is to fit two linear regression models by regressing life expectancy on infant mortality and population size separately. 
```{r}
# fit a linear model to regress life expectancy on infant mortality using only 2000 data
fit1 <- lm(life_expectancy ~ infant_mortality, data = africadata2000)
summary(fit1)

```
# Conclusion
The significant p-value (p=<0.0001) indicated a linear relationship between infant mortality and life expectancy, and the coefficient showed that the life expectancy goes down by `r round(abs(summary(fit1)$coefficient[2,1]),2)` for every 1 unit increase in infant mortality.   

```{r}
# fit a linear model to regress life expectancy on population size using only 2000 data
fit2 <- lm(life_expectancy ~ population, data = africadata2000)
summary(fit2)
```
# Conclusion
The insignificant p-value (p=`r summary(fit2)$coefficient[2,4]`) and a coefficient close to zero indicated that an absence of linear association between population size and life expectancy. 

# More Analysis (by Savannah)
I am interested in seeing a similar analysis to what we have already done with the Africa data, but this time apply it to other continents! 

I will use one additional package, `broom`, to help present my output, so I'll load that now.
```{r}
library(broom)
```


First, I want to visualize the relationship between infant mortality and life expectancy, and I'll use colors to differentiate between the continents. 

```{r}
# Create an object with just the infant mortality, life expectancy, and continent data
v1smh <- 
  select(gapminder, infant_mortality, life_expectancy, continent)

# Plot the data with different colors being different continents
ggplot(data = v1smh, aes(infant_mortality, life_expectancy, color = continent)) +
  geom_point() +
  geom_smooth()  

```

This is pretty messy, and doesn't show much, so I want to try and visualize it a different way. 

```{r}
# Plotting the data the same way, this time with different continents each having their own plots 
ggplot(data = v1smh, aes(infant_mortality, life_expectancy)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~ continent)
```
This is a little more clear, and shows the same basic trend (which we would expect) - as infant mortality increases, life expectancy decreases. One noticible difference between the continents is Africa and Asia seem to have increased infant mortality as compared to the others. 

Now I'll do the same thing with the population and life expectancy. 
```{r}
#Create another object, now with population instead of infant mortality 
v2smh <- 
  select(gapminder, population, life_expectancy, continent)

# Plot the data with different colors being different continents, with population on a log scale 
ggplot(data = v2smh, aes(population, life_expectancy, color = continent)) +
  scale_x_continuous(trans = 'log10') +
  geom_point() +
  geom_smooth()  
```

Again, this is a MESS. Let's try it again. 

```{r}
# Plot life expectancy against log population
ggplot(data = v2smh, aes(population, life_expectancy)) +
  scale_x_continuous(trans = 'log10') +
  geom_point() +
  geom_smooth() +
  facet_wrap(~ continent) +
  xlab('Population (log scale)') +
  ylab('Life Expectancy') 

```


Definitely more readable, but as with our first run-through on the analysis, there doesn't seem to be a true trend. I'm going to focus on this for an actual analysis. 

First I want to see if there are any years that are missing data, so that I can choose one year to focus on without missing data. 

```{r}
# Create object that only contains missing population information according to year and continent 
whatsmissing <- 
  gapminder %>%
  select(year, population, continent) %>%
  filter(is.na(population))
```

2016 appears to be the only year with missing population data, so for consistency's sake, I'll use the year 2000 again. 

```{r}
#Create an object containing population, life expectancy, continent, and year data, and then narrow that down to only data from the year 2000 
poplife2000 <- 
  select(gapminder, population, life_expectancy, continent, year) %>%
  filter(year == 2000)
```

Now I'll recreate the plots with the year 2000 data. 
```{r}
#Plot life expectancy against log population for the year 2000
ggplot(data = poplife2000, aes(population, life_expectancy)) +
  scale_x_continuous(trans = 'log10') +
  geom_point() +
  geom_smooth() +
  facet_wrap(~ continent) +
  xlab('Population (log scale)') +
  ylab('Life Expectancy') 

```


Less cluttered, but not super clear. I'll run a simple linear regression to see if there's any significant relationship overall. 

```{r}
# Fit a simple linear model with population and life expectancy across all continents 
fitall <- lm(life_expectancy ~ population, data = poplife2000)
summary(fitall)
tidy(fitall)
```

If we look at the relationship overall, there is _definitely_ no significant relationship here. What if we take it to individual continents?

```{r}
# Create object with just Africa from year 2000 data
afr2000 <-
  poplife2000 %>%
  filter(continent == "Africa") 

# Fit simple linear model with population and life expectancy for Africa 
fitafr <- lm(life_expectancy ~ population, data = afr2000)
summary(fitafr)
tidy(fitafr)
```

```{r}
# Create object with just Americas from year 2000 data
ame2000 <-
  poplife2000 %>%
  filter(continent == "Americas") 

# Fit simple linear model with population and life expectancy for Americas 
fitame <- lm(life_expectancy ~ population, data = ame2000)
summary(fitame)
tidy(fitame)
```

```{r}
# Create object with just Asia from year 2000 data
asi2000 <-
  poplife2000 %>%
  filter(continent == "Asia") 

# Fit simple linear model with population and life expectancy for Asia 
fitasi <- lm(life_expectancy ~ population, data = asi2000)
summary(fitasi)
tidy(fitasi)
```

```{r}
# Create object with just Europe from year 2000 data
eur2000 <-
  poplife2000 %>%
  filter(continent == "Europe") 

# Fit simple linear model with population and life expectancy for Europe 
fiteur <- lm(life_expectancy ~ population, data = eur2000)
summary(fiteur)
tidy(fiteur)
```


```{r}
# Create object with just Oceania from year 2000 data
oci2000 <-
  poplife2000 %>%
  filter(continent == "Oceania") 

# Fit simple linear model with population and life expectancy for Africa 
fitoci <- lm(life_expectancy ~ population, data = oci2000)
summary(fitoci)
tidy(fitoci)
```

Unfortunately, it doesn't look like there are relationships, at least at this level. However, I think looking for these things is important, even if you don't find any statistical relationships! And it's always fun to do quick analyses to check! 

